Modal Abstractions for Virtualizing Memory Addresses

### 2024-07-01 examples.tex
+

\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \texttt{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{\_} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta  }_{\rtv}$
mov    -0x8[rbp],rdi
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta   }_{\rtv}$
mov     rdi, r8
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry}  \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta  }_{\rtv}$
$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }\urcorner}$
mov    [r8],rdi
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry\_val} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta  }_{\rtv}$
$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }  \urcorner }$
and    0x1,rdi
mov    rdi,rax
;;  if (!entry->present) {
cmp    0x0,rax
jne    161 <pte_get_next_table_succ+0xa1>
lea    -0x10[rbp],rdi
;;pte_initialize(&next);
callq  70 <pte_initialize>
...
;;}
;; uintptr_t next_phys_addr = PTE_PFN_TO_ADDR(entry->pfn);
...
;;uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr);
mov    -0x18[rbp],rax
movabs 0x0,rcx	                187: R_X86_64_64	KERN_BASE
add    rcx,rax
...
\end{lstlisting}
\vspace{-1em}
\caption{Making sure that \textsf{entry}'s frame is referencing to an address backed up by a physical page.}
\label{fig:calltopteinitialize}
\end{figure}

@@

Consequently, before Line 7 in Figure \ref{fig:calltopteinitialize}, to read the value referenced by physical address \textsf{entry}, first we open our the addresss-space invariant ($\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)$), and then obtain the virtual mapping out of the identity mapping $\Xi$ with the side condition
\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}
 assuring that looking up identity mappings are safe, because the full ownership asserts the non-presence of the entry which requires allocation a new physical page and setting the presence bit. By obtaining the identity mapping for \textsf{entry} in Figure \ref{fig:calltopteinitialize} -- ($\textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val}$ with page addr \textsf{entry}),
 \[
 \begin{array}{l}
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{entry\},m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry} \ast \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
mov    [r8],rdi
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry\_val} \ast  \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
\end{array}
 \]
 we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit via $and 0x1,rdi$.

 \subsubsection{Ensuring Presence of Entry}
 Accessing the presence bit and checking the value allow (Lines 7-13 in Figure \ref{fig:calltopteinitialize}) us to exploit the fact in the side condition (*) in such a way that the condition ($\textsf{entry}\mapsto\textsf{present}$ determines whether we take the non-presence path (Lines 12-19 in Figure \ref{calltopteinitialize}) with full ownership ($\textsf{qfrac} = 1$) to set the entry with proper values for the bitfields, e.g., frame showing freshly allocated page and presence bit.

In case of absence, \textsf{pte\_initialize} is called for a physical-zeroed page (\textsf{kalloc} -- currently an axiom as it is not part of \textsf{pte} module but the allocator's) to be backing the physical frame address of the entry.
\mytodo{Discuss with Colin: v2p and spec on pteinitialize. After clarification expand this piece a little bit more. Right now we have physical pointsto returned from kalloc so pfn is directly set not through v2p}
Once we make sure that the frame of the page-table entry is backed with an allocated physical page, we can obtain the virtual address of the next level-page table (\textsf{next} of function \textsf{get\_next\_table} in Figure \ref{}) via looking up the \textit{identity mapping} from the physical address (the address value referenced by $\textsf{entry}\mapsto\textsf{pfn}$) to virtual address value to be returned by \textsf{get\_next\_table}.
\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
;;if (!entry->present) {
  ...

;;}
...
mov    rax,-0x18[rbp]
;;uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr);
mov    -0x18[rbp],rax
movabs $0x0,rcx            187: R_X86_64_64	KERN_BASE
add    rcx,rax
mov    rax,-0x20[rbp]
\vspace{-1em}
\caption{Converting a physical address, whose exsitence is made certain, of a PTE to a virtual address (w/o instruction pointer or flag updates). $R$ abbreviates the resources
associated with \paddr\ by $\Xi$ in Figure \ref{fig:peraspaceinvariant_with_p2v_extension}.}
\label{fig:p2v}
\end{figure}

>>>>

;;pte_t *pte_get_next_table_succ(pte_t *entry) {
push   rbp
mov    rsp,rbp
sub    0x20,rsp
mov    rdi,-0x8[rbp]
;; pte_t *next;
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \texttt{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{\_}  }_{\rtv}$
$\specline{\sumpv\paddr\qfrac\true \ast \sumapaces\rtv\delta  }_{\rtv}$
mov    -0x8[rbp],rdi
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{\_}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry}  }_{\rtv}$
$\specline{\sumpv\paddr\qfrac\true \ast \sumapaces\rtv\delta  }_{\rtv}$
mov     rdi, r8
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} }_{\rtv}$
$\specline{\sumpv\paddr\qfrac\true \ast \sumapaces\rtv\delta  }_{\rtv}$
$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }\urcorner}$
mov    [r8],rdi
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val}   }_{\rtv}$
$\specline{\sumpv\paddr\qfrac\true \ast \sumapaces\rtv\delta  }_{\rtv}$
$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }  \urcorner }$
and    0x1,rdi
mov    rdi,rax
;;  if (!entry->present) {
cmp    0x0,rax
jne    161 <pte_get_next_table_succ+0xa1>
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \ast \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val} }_{\rtv}$
$\specline{\sumpv\paddr\qfrac\true \ast \sumapaces\rtv\delta  }_{\rtv}$
$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }  \urcorner }$
mov    rbp,rdi
sub    0x10, rdi
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \ast \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val}  }_{\rtv}$
$\specline{\sumpv\paddr\qfrac\true \ast \sumapaces\rtv\delta  }_{\rtv}$
$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \land \lnot\textsf{entry\_present }  \urcorner }$
;;pte_initialize(&next);
callq  70 <pte_initialize>
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val}  }_{\rtv}$
$\specline{\sumpv\paddr\qfrac\true \ast \sumapaces\rtv\delta  }_{\rtv}$
$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \land \lnot\textsf{entry\_present }  \urcorner }$
$\specline{ \texttt{rbp-16} \mapsto_{\textsf{vpte,1}} \textsf{nextpaddr wzero}  \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{nextpaddr}  }$
;;entry->present = 1;
...
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val}  }_{\rtv}$
$\specline{\sumpv\paddr\qfrac\true \ast \sumapaces\rtv\delta  }_{\rtv}$
$\specline{ \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \land \lnot\textsf{entry\_present }  \urcorner }$
$\specline{ \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{nextpaddr}  \ast \texttt{rax} \mapsto_{\textsf{r}} \textsf{nextpaddr}  }$
;;}
...
;;uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr);
mov    -0x18[rbp],rax
movabs 0x0,rcx	                187: R_X86_64_64	KERN_BASE
add    rcx,rax
...
\vspace{-1em}
\caption{Making sure that \textsf{entry}'s frame is referencing to an address backed up by a physical page.}
\label{fig:calltopteinitialize}

@@

Consequently, before Line 7 in Figure \ref{fig:calltopteinitialize}, to read the value referenced by physical address \textsf{entry}, first we open our the addresss-space invariant ($\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)$), and then obtain the virtual mapping out of the identity mapping $\Xi$ with the side condition
\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}
 assuring that looking up identity mappings are safe, because the full ownership asserts the non-presence of the entry which requires allocation a new physical page and setting the presence bit. By obtaining the identity mapping for \textsf{entry} in Figure \ref{fig:calltopteinitialize} -- ($\textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val}$ with page addr \textsf{entry}),
 \[
 \begin{array}{l}
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus\{entry\},m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry} \ast \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
mov    [r8],rdi
\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast r8  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast rdi  \mapsto_{\textsf{r}} \textsf{entry\_val} \ast  \textsf{entry-KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry\_val} \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}
\end{array}
 \]
 we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit via $and \;0x1,rdi$.

 \subsubsection{Ensuring Presence of an Entry}
 Accessing the presence bit and checking the value allow (Lines 7-13 in Figure \ref{fig:calltopteinitialize}) us to exploit the fact in the side condition (*) in such a way that the condition ($\textsf{entry}\mapsto\textsf{present}$ determines whether we take the non-presence path (Lines 12-19 in Figure \ref{calltopteinitialize}) with full ownership ($\textsf{qfrac} = 1$) to set the entry with proper values for the bitfields, e.g., frame showing freshly allocated page and presence bit.

In case of absence, \textsf{pte\_initialize} is called for a physical-zeroed page (utilizing the page-allocator's \textsf{kalloc} -- currently the only axiom as it is not part of \textsf{pte} (virtual-memory-management) module) to be backing the physical frame address of the entry.
\mytodo{Discuss with Colin: v2p and spec on pteinitialize. After clarification expand this piece a little bit more. Right now we have physical pointsto returned from kalloc so pfn is directly set not through v2p}
Once we make sure that the frame of the page-table entry ($\textsf{entry}\mapsto\textsf{pfn}$)is backed with an allocated physical page, we can obtain the virtual address of the next level-page table. To do so, again, we refer to the  (\textsf{next} of function \textsf{get\_next\_table} in Figure \ref{}) via looking up the \textit{identity mapping} from the physical address (the address value referenced by $\textsf{entry}\mapsto\textsf{pfn}$) to virtual address value to be returned by \textsf{get\_next\_table}.
\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
;;if (!entry->present) {
  ...

;;}
...
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)  \ast \texttt{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{\_} \ast \sumpv\paddr\qfrac\true  \sumapaces\rtv\delta  }_{\rtv}$ 
mov    rax,-0x18[rbp]

;;uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr);
mov    -0x18[rbp],rax
movabs 0x0,rcx            187: R_X86_64_64	KERN_BASE
add    rcx,rax
mov    rax,-0x20[rbp]
\end{lstlisting}
\vspace{-1em}
\caption{Converting a physical address, whose exsitence is made certain, of a PTE to a virtual address (w/o instruction pointer or flag updates). $R$ abbreviates the resources
associated with \paddr\ by $\Xi$ in Figure \ref{fig:peraspaceinvariant_with_p2v_extension}.}
\label{fig:p2v}
\end{figure}
